---
title: "Exploratory Analysis"
author: "Laurel Safranek, Salma Mohamed, Gulsima Young"
output:
  prettydoc::html_pretty:
    theme: hpstr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("analysis.R")
```

## Distribution of Variables

---

One of the main variables that we are interested in focusing on for our project
is the `che_gdp` column of the Global Health Expenditure Database. This metric
stands for Current Health Expenditure (CHE) as percentage of Gross Domestic 
Product (GDP) (%). This is often referred to as the CHE%GDP. The CHE value 
explains how much a country is spending on health-related expenditures. 

When we measure this as a percent of a country's overall GDP, we are able to 
learn how much money the country is actively allocating towards health related 
funding and costs. Overall, this gives us a sense of how various countries 
prioritize public health.

```{r che_gdp_table, echo = FALSE, warning = FALSE, message = FALSE}
che_gdp_table <- GHED_data %>%
  select(country, year, che_gdp)

che_gdp_filter <- che_gdp_table %>%
  filter(country == "Algeria" | country == "Ghana" |
           country == "Canada" | country == "Haiti" | country == "France" | 
           country == "Germany" | country == "United States of America" |
           country == "Iraq" | country == "Pakistan" | country == "Australia") %>%
  filter(year == "2018")

che_gdp_df <- data.frame(country = che_gdp_filter$country, year = che_gdp_filter$year,
                         che_gdp = che_gdp_filter$che_gdp)

ggplot(data = che_gdp_df, aes(x = country, y = che_gdp)) +
   geom_bar(stat = "identity", fill = "cornflowerblue") +
   labs(
      title = "Current Health Expenditure as Percentage of GDP in 2018",
      subtitle = "",
      x = "Country",
      y = "Current Health Expenditure as Percentage of GDP"
   ) +
   theme(axis.text.x = element_text(angle = 20))
```


---

In the U.S. Health Insurance Coverage database, one of the main variables we are focusing on is the uninsurance rate.  The uninsurance rate is the percentage of residents in a certain area who are not covered by health insurance (government or private). This database spans the time period directly before and after the Affordable Care Act was implemented. Therefore analyzing the change in uninsurance rate is useful in determining the impact (if any) that the legislation had on this metric in order to guide future lawmaking. 

This visualization shows the uninsurance rate in a selection of states, as well as the United States as a whole, in 2010 (before the ACA) and 2015 (after the ACA).  Across the all regions seen here, the uninsurance rate decrease after the Affordable Care Act was put in place.  

```{r uninsured_by_state, echo = FALSE, warning = FALSE, message = FALSE}
uninsured_2010 <- as.numeric(str_trim(
                  str_remove(health_insurance_coverage$uninsured_2010, "%")))
uninsured_2015 <- as.numeric(str_trim(
  str_remove(health_insurance_coverage$uninsured_2015, "%")))

health_insurance <- data.frame(State=health_insurance_coverage$State,
                               uninsured_2010, uninsured_2015) %>%
  filter(State == "Alaska " | State == "California" | State == "Nevada" |
           State == "Massachusetts" | State == "Iowa " | State == "Hawaii" |
           State == "United States" | State == "Kentucky" | State == "Florida ") %>%
  pivot_longer(
    cols = starts_with("uninsured_"),
    names_to = "Year",
    values_to = "uninsured"
  )

ggplot(data=health_insurance, aes(x=State, y=uninsured)) +
  geom_col(aes(fill=Year), position=position_dodge(), width=0.9) +
  labs(title="Uninsurance Rate by State, Compared Across Years", subtitle = "", y="Uninsurance Rate (%)") +
  theme(axis.text.x = element_text(angle = 20)) +
  scale_fill_discrete(labels = c('2010','2015'))
```


## Relationships Between Variables

---

In order to discover relationships between variables in the context of our
project, we decided to further analyze the CHE%GDP value by comparing it with
the `year` variable. This allowed us to make predictions about how the CHE%GDP
would fluctuate, grow, or decline over time. We predicted that the CHE%GDP
would grow steadily over time, with a couple of shifts based upon various
health and economic related global events.

Though this graph does show a slight upward trend over time, it was not nearly
as drastic as we thought. However, this is still vital information to have
as we continue to dissect and examine our data in order to search for trends
and answers to the questions that we have.

```{r year_vs_che_gdp_table, echo = FALSE, warning = FALSE, message = FALSE}
year_vs_che_gdp_table <- GHED_data %>%
  select(country, year, che_gdp)

che_gdp_filter <- year_vs_che_gdp_table %>%
  filter(country == "Algeria" | country == "Ghana" |
           country == "Canada" | country == "Haiti" | country == "France" | 
           country == "Germany" | country == "United States of America" |
           country == "Iraq" | country == "Pakistan" | country == "Australia")

che_gdp_year_df <- data.frame(country = che_gdp_filter$country, 
                              year = che_gdp_filter$year, che_gdp = 
                                che_gdp_filter$che_gdp)

ggplot(che_gdp_year_df, aes(year, che_gdp, color = country)) +
   geom_point(mapping = aes(x = year, y = che_gdp, color = country)) +
   labs(
      title = "Current Health Expenditure as Percentage of GDP Over Time",
      subtitle = "",
      x = "Country",
      y = "Current Health Expenditure as Percentage of GDP"
   ) +
   theme(axis.text.x = element_text(angle = 20))
```


---
As part of the Affordable Care Act, the eligibility criteria for Medicaid were expanded to include more people.  Each state could decide whether to expand Medicaid for its residents, with the federal government supplying some to all of the funding for this expansion for a period of a several years.  Ultimately, some states expanded Medicaid, while other states did not.  

Here, we compare the percentage of residents without health insurance coverage in states that did expand Medicaid and states that did not.  We can see that in states which opted to increase Medicaid eligibility criteria, less people ended up being uninsured, possibly because they were able to be covered by Medicaid.

```{r medicaid_expansion_chart, echo = FALSE, warning = FALSE, message = FALSE}
medicaid <- health_insurance_coverage$expansion
uninsured_2015_stripped <- as.numeric(str_trim(
  str_remove(health_insurance_coverage$uninsured_2015, "%")))

df <- data.frame(uninsured_2015_stripped, medicaid) %>%
  filter(medicaid == "True" | medicaid == "False") %>%
  group_by(medicaid) %>%
  summarise(uninsured=mean(uninsured_2015_stripped))


ggplot(data=df, aes(x=uninsured, y=medicaid)) +
  geom_col(fill="cornflowerblue", width=0.5) +
  labs(title="Medicaid Expansion vs. Percent Uninsured",
       y="Medicaid Expansion", x="Uninsured in 2015 (%)")

```

